---
title: "Structural analysis of first-price auctions with qdfauction: the USFS timber data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural analysis of first-price auctions with qdfauction: the USFS timber data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, collapse = TRUE, comment = "#>")
set.seed(20260925)
```

This vignette walks through the full pipeline of `qdfauction` on real data:
the sealed-bid subsample of the U.S. Forest Service timber auctions assembled
by Lu and Perrigne (2008). We recover bidders' private values from their bids,
estimate the private-value density, and estimate the seller's expected revenue
as a function of the reserve price.

```{r}
library(qdfauction)
data(timber)
str(timber)
```

The data contain 107 auctions with two bidders and 108 with three. We analyse
the two groups separately so that the number of bidders is known and fixed
within each group.

## 1. The quantile density of bids

In a symmetric first-price auction with independent private values the
inverse bid function is (Guerre, Perrigne and Vuong 2000)

$$v = b + \frac{1}{n-1}\frac{G(b)}{g(b)} = b + \frac{u\,q(u)}{n-1},\qquad u = G(b),$$

where $q = 1/g(Q(u))$ is the quantile density of bids. The second form
(Zhang 2022; Doosti, Dewan and Talebian 2025) avoids dividing by a density
estimate that is small and unstable near the boundaries. `qdf()` estimates
$q$ with ten estimators; the indirect Poisson estimator with biased
cross-validation performed best in the 2025 simulations.

```{r}
b3  <- as.numeric(timber$n3)
# bandwidth = "bcv" selects h by biased cross-validation (about 20 s here);
# we fix h at the value it selects to keep the vignette fast
q_ip <- qdf(b3, method = "indirect_poisson", bandwidth = 0.0996)
q_b  <- qdf(b3, method = "bernstein",        bandwidth = "bcv")
q_w  <- qdf(b3, method = "wavelet_block",    bandwidth = 0.15)
q_ip
plot(q_ip, ylim = c(0, 400), main = "Quantile density of bids, N = 3")
lines(q_b$u, q_b$q, col = 2); lines(q_w$u, q_w$q, col = 4)
legend("topleft", c("indirect Poisson", "Bernstein", "block wavelet"),
       col = c(1, 2, 4), lty = 1, bty = "n")
```

The quantile density rises steeply toward $u = 1$: the bid density is close
to zero at the top of the support, which is exactly where the conventional
ratio $G/g$ is unstable.

## 2. Recovering private values

`fpa_values()` applies the inversion. We compare the indirect Poisson
inversion with the conventional GPV estimator and with the boundary-corrected
kernel of Hickman and Hubbard (2015), and impose monotonicity of the inverse
bid function with the natural-spline fix of the 2025 paper.

```{r}
fv_ip  <- fpa_values(timber$n3, method = "indirect_poisson", bandwidth = q_ip$h,
                     monotone = "spline")           # reuse the BCV bandwidth from above
fv_gpv <- fpa_values(timber$n3, method = "gpv")
fv_ip
plot(fv_ip, main = "Inverse bid function, N = 3")
points(fv_gpv$bids, fv_gpv$values, col = 2, pch = 3, cex = 0.6)
legend("topleft", c("indirect Poisson (raw)", "monotone fit", "GPV"),
       col = c("grey50", 4, 2), pch = c(20, NA, 3), lty = c(NA, 1, NA), bty = "n")
```

The implied markups $\hat v - b$ fall with competition, as theory predicts:

```{r}
fv2 <- fpa_values(timber$n2, method = "indirect_poisson", bandwidth = 0.0975,   # BCV choice
                  monotone = "spline")
rbind(`N = 2` = quantile(fv2$values - fv2$bids, c(.25, .5, .75)),
      `N = 3` = quantile(fv_ip$values - fv_ip$bids, c(.25, .5, .75)))
```

Lu and Perrigne (2008) find evidence of risk aversion in these auctions.
Under CRRA utility $U(t) = t^{1-\eta}$ the inversion becomes
$v = b + (1-\eta)\,u q(u)/(n-1)$, so risk-neutral pseudo-values are upper
bounds on the true ones. The `eta` argument handles this:

```{r}
fv_ra <- fpa_values(timber$n3, method = "indirect_poisson", bandwidth = q_ip$h,
                    eta = 0.3, monotone = "spline")
median(fv_ra$values - fv_ra$bids) / median(fv_ip$values - fv_ip$bids)   # = 1 - eta
```

## 3. The private-value density

`fpa_density()` estimates the density of private values in two steps. The
first step recovers pseudo-values (Hickman–Hubbard boundary-corrected GPV
by default, or any `fpa_values()` fit); the second step estimates their
density with the wavelet estimators of Doosti, Dewan and Sbaï (2026) or one
of the kernel competitors. Because the bids are in dollars, they are mapped
to $[0,1]$ by a log map and the density is returned in dollars through the
Jacobian.

```{r}
fd_w  <- fpa_density(timber$n3, method = "wavelet_ll", j0 = 4, h = 0.15)
fd_hh <- fpa_density(timber$n3, method = "kde_hh")
fd_ms <- fpa_density(timber$n3, method = "marmer_shneyerov")
fd_q  <- fpa_density(timber$n3, method = "wavelet_ll", first_stage = fv_ip, j0 = 4, h = 0.15)
fd_w
plot(fd_w, xlim = c(0, 400), ylim = c(0, 0.02), main = "Private-value density, N = 3")
lines(fd_hh$x, fd_hh$f, col = 2, lty = 2)
lines(fd_ms$x, fd_ms$f, col = 3, lty = 3)
lines(fd_q$x,  fd_q$f,  col = 4, lty = 4)
legend("topright", c("wavelet LL (GPV-HH first stage)", "GPV-HH kernel",
                     "Marmer-Shneyerov", "wavelet LL (QDF first stage)"),
       col = 1:4, lty = 1:4, bty = "n")
```

The tuning parameters of the wavelet estimator can be chosen by
cross-validated held-out log-likelihood (`j0 = "cv"`), and the estimators
can be compared on the same footing with `fpa_density_cv()`:

```{r, eval = FALSE}
fd_cv <- fpa_density(timber$n3, method = "wavelet_ll", j0 = "cv")
fpa_density_cv(timber$n3, methods = c("wavelet_ll", "kde_hh", "marmer_shneyerov"),
               folds = 5, j0 = fd_cv$j0, h = fd_cv$h)
```

## 4. Expected revenue and the optimal reserve price

Given recovered values, `fpa_revenue()` estimates the seller's expected
revenue as a function of the reserve price (Zincenko 2024; Doosti 2026b), and
`fpa_optimal_reserve()` maximises it. Here we take the seller's own valuation
to be `v_s = 20` dollars per unit.

```{r}
r  <- seq(20, 120, by = 5)
R3 <- fpa_revenue(fv_ip, r, v_s = 20)
opt <- fpa_optimal_reserve(fv_ip, v_s = 20)
plot(r, R3, type = "l", lwd = 2, xlab = "reserve price ($/unit)",
     ylab = "expected revenue", main = "Expected revenue, N = 3")
abline(v = opt$r_star, lty = 2)
opt[c("r_star", "revenue")]
```

Bootstrap standard errors, pointwise intervals and a uniform band follow
from `fpa_revenue_boot()`; the Bernstein estimator is used here because it is
fast enough for many bootstrap replications.

```{r}
rb <- fpa_revenue_boot(timber$n3, r = seq(30, 100, by = 10), B = 30, v_s = 20,
                       method = "bernstein", bandwidth = "wbcv", m_range = c(10, 200))
rb
plot(rb, main = "Expected revenue with 95% pointwise CI and uniform band")
```

## 5. Checking the method by simulation

`fpa_simulate()` draws values from any marginal (and, optionally, an
Archimedean copula for affiliated values or a CRRA risk-aversion parameter)
and computes exact equilibrium bids, and `fpa_montecarlo()` scores a list of
estimators over replications. The following reproduces one row of Table 2 of
Doosti, Dewan and Talebian (2025) in a few seconds:

```{r}
pdf <- function(x) dgamma(x, 10, 2); cdf <- function(x) pgamma(x, 10, 2)
mc <- fpa_montecarlo(R = 3, L = 20, n = 5, Q = function(u) qgamma(u, 10, 2), seed = 1,
  estimators = list(
    B_B  = function(b, v) fpa_values(b, method = "bernstein",        bandwidth = "bcv"),
    B_IP = function(b, v) fpa_values(b, method = "indirect_poisson", bandwidth = "bcv"),
    Conv = function(b, v) fpa_values(b, method = "gpv")),
  score = function(fit, b, v) fpa_evaluate_values(fit, pdf, cdf, v))
aggregate(cbind(mise, loglik, ks) ~ estimator, mc, mean)
```

## References

Doosti, H. (2026a). A copula-quantile density approach to affiliated private value auctions. *Economics Letters*, 261, 112837.

Doosti, H. (2026b). Estimating seller's expected revenue in first-price sealed-bid auctions via quantile density functions. Working paper.

Doosti, H., Dewan, I. and Sbaï, E. (2026). Adaptive wavelet estimation of private-value densities in first-price auctions. Working paper.

Doosti, H., Dewan, I. and Talebian, M. (2025). Nonparametric estimation of private value distributions in first-price auctions: Evaluating quantile density function approaches. *Economics Letters*, 257, 112670.

Guerre, E., Perrigne, I. and Vuong, Q. (2000). Optimal nonparametric estimation of first-price auctions. *Econometrica*, 68, 525–574.

Hickman, B. R. and Hubbard, T. P. (2015). Replacing sample trimming with boundary correction in nonparametric estimation of first-price auctions. *Journal of Applied Econometrics*, 30, 739–762.

Lu, J. and Perrigne, I. (2008). Estimating risk aversion from ascending and sealed-bid auctions: the case of timber auction data. *Journal of Applied Econometrics*, 23, 871–896.

Zincenko, F. (2024). Estimation and inference of seller's expected revenue in first-price auctions. *Journal of Econometrics*, 241, 105734.
